---
title: "CTmax of bopyrid infested *Acartia tonsa*"
date: "`r Sys.Date()`"
output: 
  html_document:
          code_folding: hide
          code_download: true
          toc: true
          toc_float: true
  github_document:
          html_preview: false
          toc: true
          toc_depth: 3
---

# Project Description
We collected a plankton sample from Key Largo Florida in late February 2023. Water temperature was 27 degrees C at the time of collection. While isolating *Acartia tonsa* individuals from this sample, we noticed that many females had larvae of a bopyrid isopod attached to their prosome. Given the important role *Acartia* copepods play in coastal systems, we wanted to test the hypothesis that infestation by bopyrid isopod larvae would reduce thermal limits in *Acartia tonsa*. 

# Methods 
After collection, *Acartia tonsa* individuals were sorted from the bulk contents of the plankton tow. As copepods were being isolated, mature females that carried a late stage bopyrid isopod larva on their prosome were set aside, with an equal number of females not carrying an isopod. When 10 individuals had been collected (5 each, with and without isopods), critical thermal maxima (CTmax) were measured for the individuals using a custom heating device.

Briefly, individuals were isolated in 10 mL of filtered seawater in glass tubes. These tubes were held in a plexiglass water bath, held above a 20 L reservoir. The reservoir contained a 300 watt aquarium heater, which raised the temperature of the water in the reservoir at a fairly constant rate of 0.1-0.3 degrees C per minute. An aquarium pump in the reservoir flooded the water bath, which resulted in gradual warming of the water in the tubes. Copepods were continuously monitored until they ceased to respond to gentle stimulus (water movement caused by rotating the tube). 

The CTmax values for the two groups (infected and non-infected) were compared using effect size estimates (mean difference). Confidence intervals were estimated using non-parametric bootstrapping. Since multiple replicate experiments were run and CTmax may have changed as time since collection increased, we also ran a linear mixed effects model, with CTmax modeled as a function of infection status with experimental replicate as a random effect. 

```{r setup, include=T, message = F, warning = F, echo = F}
knitr::opts_chunk$set(
  echo = knitr::is_html_output(),
  fig.align = "center",
  fig.path = "../Figures/markdown/",
  dev = c("png", "pdf"),
  message = FALSE,
  warning = FALSE,
  collapse = T
)

theme_matt = function(base_size = 18,
                      dark_text = "grey20"){
  mid_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  ggpubr::theme_pubr(base_family="sans") %+replace% 
    theme(
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title = element_text(size = base_size * 1.2,
                                margin = margin(0, 8, 0, 0)),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}
```

# Results
There was no difference between CTmax values of copepods infested with isopods and the non-infested individuals. The estimated effect size (mean difference) was positive, but the confidence interval strongly overlapped zero. 

```{r eff-size}
bop_eff = dabest(full_data,
                 x = bopyrid,
                 y = ctmax, 
                 idx = c("no", "yes"),
                 paired = F)

bop_diff = mean_diff(bop_eff)

plot(bop_diff)
```

A linear mixed effect model also indicated no effect of bopyrid infestation. This model examined CTmax as a function of infestation status, with run number as a random effect. 

```{r lme-model}
iso.model = nlme::lme(data = full_data, 
                      fixed = ctmax ~ bopyrid, random = ~1|run)
kable(car::Anova(iso.model))
```

One interesting aspect of this data is that while bopyrid infestation did not result in changes in the mean CTmax value, there was an apparent increase in the variability of thermal limits between groups. Within the isopod infested individuals, there were three substantially lower thermal limits. When these are removed, there is a small effect size estimate (bopyrid infested individuals had higher thermal limits than the controls).  

```{r subset-eff-size}
subset = full_data %>%  
  filter(ctmax > 36)  

bop_eff = dabest(subset,
                 x = bopyrid,
                 y = ctmax, 
                 idx = c("no", "yes"),
                 paired = F)

bop_diff = mean_diff(bop_eff)

plot(bop_diff)
```

With the three low values removed, a linear mixed effects model also indicates an effect of bopyrid infestation on CTmax. 

```{r sub-lme-model}
sub_iso.model = nlme::lme(data = subset, 
                      fixed = ctmax ~ bopyrid, random = ~1|run)
kable(car::Anova(sub_iso.model))
```

# Discussion
There was no effect of bopyrid infestation on the thermal limit of *Acartia tonsa* females. If several notably lower thermal limits were excluded, however, thermal limits appear to be higher in the infested copepods than in the non-infested individuals. In both cases (no difference or a small increase in thermal limits), this is contrary to our expectations that bopyrid infestation would reduce copepod thermal limits. 

The pattern observed is strongly affected by whether we include or exclude three notably lower thermal limit measurements in the infested treatment. The attached bopyrids varied in size within this treatment. Assuming individuals are not switching hosts, this means that females with larger bopyrids are likely older than females with smaller bopyrids. The three individuals with particularly low thermal limits had larger bopyrids; the low thermal limits may therefore reflect the older age of these females. However, unfortunately without proper equipment we were unable to take photographs of these individuals and their bopyrids in order to measure lengths. 

There are several potential explanations for an increase in thermal limits when parasitized. It's possible that HSP production is stimulated by infestation, thus priming copepods to respond to heat stress. Alternatively, previous work shows reduced respiration rates in *Acartia* infested by bopyrids. Reduced basal metabolisms may improve thermal tolerance. Reduced metabolism may also hinder the thermal stress response though if ATP limitation affects the production of HSPs.

# Other things to add
Depending on the number of similar studies, it might be possible to add a small meta-analysis on the effects of parasitism on CTmax. 

