---
title: "CTmax of bopyrid infested *Acartia tonsa*"
date: "`r Sys.Date()`"
output: 
  html_document:
          code_folding: hide
          code_download: true
          toc: true
          toc_float: true
  github_document:
          html_preview: false
          toc: true
          toc_depth: 3
---

# Project Description
We collected a plankton sample from Key Largo Florida in late February 2023. Water temperature was 27 degrees C at the time of collection. While isolating *Acartia tonsa* individuals from this sample, we noticed that many females had larvae of a bopyrid isopod attached to their prosome. Given the important role *Acartia* copepods play in coastal systems, we wanted to test the hypothesis that infestation by bopyrid isopod larvae would reduce thermal limits in *Acartia tonsa*. 

# Methods 
After collection, *Acartia tonsa* individuals were sorted from the bulk contents of the plankton tow. As copepods were being isolated, mature females that carried a late stage bopyrid isopod larva on their prosome were set aside, with an equal number of females not carrying an isopod. When 10 individuals had been collected (5 each, with and without isopods), critical thermal maxima (CTmax) were measured for the individuals using a custom heating device.

Briefly, individuals were isolated in 10 mL of filtered seawater in glass tubes. These tubes were held in a plexiglass water bath, held above a 20 L reservoir. The reservoir contained a 300 watt aquarium heater, which raised the temperature of the water in the reservoir at a fairly constant rate of 0.1-0.3 degrees C per minute. An aquarium pump in the reservoir flooded the water bath, which resulted in gradual warming of the water in the tubes. Copepods were continuously monitored until they ceased to respond to gentle stimulus (water movement caused by rotating the tube). 

```{r setup, include=T, message = F, warning = F, echo = F}
knitr::opts_chunk$set(
  echo = knitr::is_html_output(),
  fig.align = "center",
  fig.path = "../Figures/markdown/",
  dev = c("png", "pdf"),
  message = FALSE,
  warning = FALSE,
  collapse = T
)

theme_matt = function(base_size = 18,
                      dark_text = "grey20"){
  mid_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  ggpubr::theme_pubr(base_family="sans") %+replace% 
    theme(
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title = element_text(size = base_size * 1.2,
                                margin = margin(0, 8, 0, 0)),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}
```

# Results
There was no difference between CTmax values of copepods infested with isopods and the non-infested individuals. 

```{r}
ggplot(full_data, aes(x = bopyrid, y = ctmax)) + 
  geom_boxplot() + 
  geom_point(position = position_jitter(width = 0.1, height = 0)) + 
  theme_matt()
```

The estimated effect size (mean difference) was positive, but the confidence interval strongly overlapped zero. 

```{r}
bop_eff = dabest(full_data,
                 x = bopyrid,
                 y = ctmax, 
                 idx = c("no", "yes"),
                 paired = F)

bop_diff = mean_diff(bop_eff)

plot(bop_diff)
```

A linear mixed effect model also indicated no effect of bopyrid infestation. This model examined CTmax as a function of infestation status, with run number as a random effect. 

```{r}
iso.model = nlme::lme(data = full_data, 
                      fixed = ctmax ~ bopyrid, random = ~1|run)
kable(car::Anova(iso.model))
```

One interesting aspect of this data is that while bopyrid infestation did not result in changes in the mean CTmax value, there was an apparent increase in the variability of thermal limits between groups. Within the isopod infested individuals, there were three substantially lower thermal limits. When these are removed, there is a large effect size estimate (bopyrid infested individuals had higher thermal limits than the controls).  

```{r}
subset = full_data %>%  
  filter(ctmax > 36)  

bop_eff = dabest(subset,
                 x = bopyrid,
                 y = ctmax, 
                 idx = c("no", "yes"),
                 paired = F)

bop_diff = mean_diff(bop_eff)

plot(bop_diff)
```

# Discussion
There was no effect of bopyrid infestation on the thermal limit of *Acartia tonsa* females. If several notably lower thermal limits were excluded, however, thermal limits appear to be higher in the infested copepods than in the non-infested individuals. In both cases, this is contrary to our expectations - that bopyrid infestation would reduce copepod thermal limits. 

There are several potential explanations for an increase in thermal limits when parasitized. It's possible that HSP production is stimulated by infestation, thus priming copepods to respond to heat stress. Alternatively, previous work shows reduced respiration rates in *Acartia* infested by bopyrids. Reduced basal metabolisms may improve thermal tolerance. 

Our determination of whether infestation affects thermal limits or not seems to depend on the three low thermal limit measurements made for individuals carrying bopyrid larvae. Bopyrid larvae grow while attached to their copepod hosts, and the isopod larvae attached to the individuals with low thermal limits were noticeably larger than average, although we lacked the ability to photograph the copepod and isopod in order to ascertain exact lengths. However, assuming the bopyrid larvae are not frequently switching hosts, copepod bearing very large larvae are presumably significantly older than adults bearing smaller isopod larvae. This may be driving the lower thermal limits of those three individuals. 

